---
params: 
  current_continent: Europe
  year: 2009
  data_file: hiv_wb.xlsx
title: "**`r paste0('Report on HIV prevalence in ', params$current_continent, ' - ', params$year)`**"
author: Sonia Mazzi - Data Science Campus
description: "An example of a reproducible report, from data importing, tidying, and cleaning, analysing to publishing with rmarkdown with govdown style. Note author is ignored in government publications."
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  govdown::govdown_document:
    title: "Report series on HIV prevalence"
    logo: "dsclogo.svg"
    logo_text: "" 
    feedback_url: "https://github.com/datasciencecampus/ARP/issues" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r}
## Required libraries
library(readxl)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(purrr)
library(DT)
library(stringr)
library(maps)
```


```{r}
#Now we will read in all worksheets and combine them into a single table, stacking the tables for each year.
#excel workbook file name is fn
#fn = "hiv_wb.xlsx"
fn <- params$data_file
combined_data <- 
  excel_sheets(fn) %>% 
  map_df(~ {read_excel(fn, sheet = .x, skip = 4, trim_ws = TRUE) %>% 
      mutate(year = as.integer(.x))}) %>% select(country, year, everything())
```

```{r}
#add geographical data
fng <- "DataGeographiesGapminder.xlsx"
geo_dat <- read_excel(fng, sheet = 2)
geo_data <- geo_dat %>% select(country = name, geo, four_regions, cen_Latitude = Latitude, cen_Longitude = Longitude)
hiv_full <- inner_join(combined_data, geo_data)
hiv_full <- hiv_full %>% select(country, geo, continent = four_regions, cen_Latitude, cen_Longitude, everything())
```

```{r}
#add percentual change
  hiv_full <- hiv_full  %>% group_by(country) %>% 
  arrange(year, .by_group = TRUE) %>%
  mutate(pct_chg_prevalence = ifelse( lag(hiv_prevalence)!= 0, ((hiv_prevalence/lag(hiv_prevalence) - 1) * 100), NA))
```

```{r}
#change some country names for maps

old_names <- c("Congo, Dem. Rep.", "Cote d'Ivoire", "Congo, Rep.", "Holy See", "Slovak Republic", "Macedonia, FYR", "United Kingdom",
               "Antigua and Barbuda", "St. Kitts and Nevis", "St. Lucia", "St. Vincent and the Grenadines", "Trinidad and Tobago",
               "United States", "Kyrgyz Republic", "Lao", "Micronesia, Fed. Sts.")
new_names <- c("Democratic Republic of the Congo",  "Ivory Coast", "Republic of Congo", "Vatican", "Slovakia", "Macedonia", "UK",
               "Antigua", "Saint Kitts", "Saint Lucia", "Saint Vincent", "Trinidad", "USA", "Kyrgyzstan", "Laos", "Micronesia")

for (i in 1:length(old_names)){
 hiv_full <- hiv_full %>% 
   ungroup() %>%
   mutate(country = str_replace(country, old_names[i], new_names[i]))
 }
```

```{r}
#capitalise the first letter of the continent names
hiv_full <- hiv_full %>% 
  mutate(continent = str_replace(continent, "africa", "Africa")) %>%
  mutate(continent = str_replace(continent, "europe", "Europe")) %>%
  mutate(continent = str_replace(continent, "asia", "Asia")) %>%
  mutate(continent = str_replace(continent, "americas", "Americas"))
```

```{r}
#Add Western Sahara
western_sahara <- hiv_full %>% 
  filter(country == "Libya") %>%
  mutate(geo = str_replace(geo, "lby", "esh")) %>%
  mutate(income = NA) %>%
  mutate(population = NA) %>%
  mutate(cen_Latitude = NA) %>%
  mutate(cen_longitude = NA) %>%
  mutate(country = str_replace(country, "Libya", "Western Sahara"))

hiv_full <- bind_rows(hiv_full, western_sahara)
```

```{r}
#add Kosovo
kosovo <- hiv_full %>% 
  filter(country == "Libya") %>%
  mutate(geo = str_replace(geo, "lby", "xkx")) %>%
  mutate(continent = str_replace(continent, "Africa", "Europe")) %>% 
  mutate(income = NA) %>%
  mutate(population = NA) %>%
  mutate(cen_Latitude = NA) %>%
  mutate(cen_longitude = NA) %>%
  mutate(country = str_replace(country, "Libya", "Kosovo"))

hiv_full <- bind_rows(hiv_full, kosovo)
```


```{r}
latest_year <- params$year
current_continent <- params$current_continent
```


# Introduction

Prevalence is a measure of disease that allows us to determine a person's likelihood of having a disease. The number of prevalent cases is the total number of cases of disease existing in a population. A prevalence rate is the total number of cases of a disease existing in a population divided by the total population. So, if a measurement of cancer is taken in a population of 40,000 people and 1,200 were recently diagnosed with cancer and 3,500 are living with cancer, then the prevalence of cancer is 0.1175. (11.75\% or 11,750 per 100,000 persons)

Using data from Gapminder we look at the latest figures on HIV prevalence in `r params$continent`.

# The Data

The data was obtained from Gapminder.org and consists of 
estimated yearly HIV Prevalence, in percentages, for people aged 15-49.

The data from 1990 onwards is from UNAIDS, but Gapminder has provided extra estimates for some countries before 1990.
Data for Brazil is the average of the high and low estimate from UNAIDS from 1990 onwards, and 1979 is the estimate from Gapminder Foundation.		



# Plots


## Plot of HIV prevalence vs GDP per capita for year `r latest_year`.

This is a plot of HIV prevalence vs GDP per capita (ppp-adjusted).

```{r}
#lower bounds for high levels of hiv
hi_hiv <- ifelse(current_continent == "Africa", 10,
                 ifelse(current_continent == "Europe", 0.5, 
                        ifelse(current_continent == "Americas", 1, 
                         ifelse(current_continent == "Asia", 0.5, NA))))
```       


```{r fig.align="center", fig.width=9, fig.height=5, fig.cap="Figure caption"}
hlhiv <- hiv_full %>% filter(continent == current_continent, year == latest_year, hiv_prevalence >= hi_hiv)
hiv_full %>% 
  filter(continent == current_continent, year == latest_year) %>% 
  ggplot(aes(x = income, y = hiv_prevalence)) +
  geom_point(aes(size = population), alpha = 0.5) +
  geom_text_repel(data = hlhiv , aes(label = country, color = "black"), col = "black", size = 3, segment.alpha = 0.8) +
  labs(x = "GDP per capita (PPP adjusted, log10)", y = "HIV prevalence (%)") +
  scale_x_log10() +
  theme(axis.text.x = element_text(color = "black", size = 5, angle = 0)) 
```



## A time plot of the  percentual change of HIV prevalence since 2000, for countries with an exceptionally high level of HIV prevalence


We consider countries with elevated HIV prevalence ( $\ge$ `r hi_hiv`) relative to the other countries in the continent.

Any point below the horizontal line at zero indicate that HIV prevalence has decreased from the previous year. Points at the horizontal zero-line indicate that HIV prevalence hasn't changed from the previous year. Points above the horizontal zero-line indicate a growth in HIV prevalence from the previous year.  

```{r warning = FALSE}
datos <- hiv_full %>% 
  filter(hiv_prevalence >= hi_hiv, year >= 2000, continent == current_continent)
#
hiv_pct_chg_plot <- ggplot(datos, aes(x = year, y = pct_chg_prevalence, color = country)) +
  geom_line() +
  geom_point(aes(size = hiv_prevalence), alpha = 0.2) +
  labs(x = "Year", y = "% change (from previous year) in HIV prevalence") +
  scale_x_continuous(limits = c(1999, latest_year), breaks = 2000:latest_year) +
  geom_text_repel(data = filter(datos, year==2000) , aes(label = geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x = -1.2) +       
  guides(size = guide_legend(title = "HIV prevalence", order = 1)) 
```


```{r fig.align="center", fig.width=9, fig.height=5, fig.cap="Figure caption"}
if(current_continent == "Africa"){
  print(hiv_pct_chg_plot + 
          geom_text_repel(data = filter(datos, year == 2002, country == "Mozambique"), 
                          aes(label = geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x = -1.2))
  }
```

```{r fig.align="center", fig.width=9, fig.height=5, fig.cap="Figure caption"}
if(current_continent == "Americas"){
  print(hiv_pct_chg_plot + 
          geom_text_repel(data = filter(datos, year == 2006, country == "Barbados"), 
                          aes(label = geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x = -1.2))
  }
```

```{r fig.align="center", fig.width=9, fig.height=5, fig.cap="Figure caption"}
if(current_continent == "Asia"){
  print(hiv_pct_chg_plot + 
          geom_text_repel(data = filter(datos, year == 2001, country == "Papua New Guinea") , aes(label = geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x=-1.2) +
          geom_text_repel(data = filter(datos, year == 2005, country == "Malaysia"), 
                          aes(label=geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x = -1.2))
  }
```

```{r fig.align="center", fig.width=9, fig.height=5, fig.cap="Figure caption"}
if(current_continent == "Europe"){
  print(hiv_pct_chg_plot + 
          geom_text_repel(data = filter(datos, year == 2001, country %in% c("Russia", "Portugal")), 
                          aes(label = geo), col = "black", size = 3, segment.alpha = 0.7, direction = "y", nudge_x = -1.2))
  }
```


## A map of `r current_continent` with countries coloured according to HIV prevalence in `r latest_year`.

```{r}
max_long <- ifelse(current_continent == "Americas", 0, 
                   ifelse(current_continent == "Asia", 185, Inf))
```

```{r}
min_long <- ifelse(current_continent == "Americas", -165, 
                   ifelse(current_continent == "Asia", 20, -Inf))
```

```{r}
min_lat <- ifelse(current_continent == "Asia", -45, -Inf)
```


```{r fig.align="center", fig.width=8, fig.height=8, fig.cap="Figure caption"}
hiv_latest_year <- hiv_full %>% 
  filter(year == latest_year, continent == current_continent)
#
current_continent_countries <- hiv_latest_year %>% 
  pull(country)
#
max_hiv <- max(hiv_latest_year$hiv_prevalence)
#
hiv_data <- map_data("world") %>% 
  filter(long >= min_long, long <= max_long, lat >= min_lat) %>%
  filter(region %in% current_continent_countries) %>% 
  mutate(country = region) %>% 
  left_join(hiv_latest_year)
#
hiv_geos <- hiv_data %>% 
  filter(long >= min_long, long <= max_long, lat >= min_lat) %>%
  select(geo, cen_Latitude, cen_Longitude) %>% 
  group_by(geo) %>% 
  unique()
#
  ggplot(hiv_data) + 
  geom_polygon(aes(x = long, y = lat, fill = hiv_prevalence, group = group, color = "")) + 
  coord_quickmap() +
  guides(fill=guide_legend(title = "HIV prevalence", order = 1)) +
  scale_fill_gradient2(low = "brown3", mid = "cornsilk1", high = "turquoise4", limits = c(0,max_hiv), na.value = "grey") +
  scale_colour_manual(values = NA) +              
  guides(colour=guide_legend("No data", override.aes = list(fill = "grey"), order = 2)) +
  geom_text_repel(data = hiv_geos, aes(x = cen_Longitude, y = cen_Latitude, label = geo), size = 3, direction = "x", segment.alpha = 0.2) +   
  theme_void()
```



```{r}
#plot Europe without Russia
hiv_latest_year <- hiv_full %>% 
  filter(year == latest_year, continent == current_continent, country != "Russia")
#
max_hiv <- max(hiv_latest_year$hiv_prevalence)
#
current_continent_countries <- hiv_latest_year %>% pull(country)
#
max_lat <- 72
#
datos_map <- map_data("world") %>% 
  filter(region %in% current_continent_countries, lat <= max_lat) %>% 
  mutate(country = region) %>% 
  left_join(hiv_latest_year)

#
hiv_geos <- datos_map %>% 
  select(geo, lat, long) %>% 
  group_by(geo) %>% 
  summarise(mean_lat = mean(range(lat), trim=0.4), mean_long = mean(range(long), trim = 0.4))
#

plot_western_europe <- ggplot(data = datos_map) + 
  geom_polygon(aes(x = long, y = lat, fill = hiv_prevalence, group = group, color = "")) + 
  coord_quickmap() +
  scale_fill_gradient2(low = "brown3", mid = "cornsilk1", high = "turquoise4", limits = c(0, max_hiv), na.value = "grey") +
  scale_colour_manual(values = NA) +              
  guides(colour=guide_legend("No data", override.aes = list(fill = "grey"), order = 2)) +
  guides(fill = guide_legend(title = "HIV prevalence", order = 1)) +
  geom_text_repel(data = hiv_geos, aes(x = mean_long, y = mean_lat, label = geo), size = 3, direction = "x", segment.alpha = 0.2) +  
  theme_void() + 
  ggtitle("Europe without Russia")  
```

```{r fig.align = "center", fig.width = 7, fig.height = 7, fig.cap = "Figure caption"}
if(current_continent == "Europe"){print(plot_western_europe)} 
```


```{r}
#The Caribbean
hiv_latest_year = hiv_full %>% filter(year == latest_year, continent == current_continent)
current_continent_countries = hiv_latest_year %>% pull(country)
#
cont <- map_data("world") %>% 
  filter(region %in% current_continent_countries, long <= -50, long >= -95, lat >= 0, lat <= 25) %>% 
  mutate(country = region) %>% 
  left_join(hiv_latest_year)
#
hiv_geos <- cont %>% 
  select(geo, lat, long) %>% 
  group_by(geo) %>% 
  summarise(mean_lat = mean(range(lat), trim = 0.4), mean_long = mean(range(long), trim = 0.4))

#
plot_caribbean <- ggplot(data = cont) + 
  geom_polygon(aes(x = long, y = lat, fill = hiv_prevalence, group = group, color = "")) + 
  coord_quickmap() +
  scale_fill_gradient2(low = "brown3", mid = "cornsilk1", high = "turquoise4", limits = c(0, 3), na.value = "grey") +
  scale_colour_manual(values = NA) +              
  guides(colour = guide_legend("No data", override.aes = list(fill = "grey"), order = 2)) +
  guides(fill = guide_legend(title = "HIV prevalence", order = 1)) +
  geom_text_repel(data = hiv_geos, aes(x = mean_long, y = mean_lat, label = geo), size = 3, direction = "x", segment.alpha = 0.2) +
  theme_void() +
  ggtitle("HIV prevalence in Central America and the Caribbean region")
```

```{r fig.align = "center", fig.width = 8, fig.height = 8, fig.cap = "Figure caption"}
if(current_continent == "Americas"){print(plot_caribbean)} 
```

# Tables

Below is a table of HIV prevalence by country from 2010 until present. The table is searchable by country name.

```{r}
table1 <- hiv_full %>% 
  filter(year >= 2000, year <= params$year, continent == params$current_continent) %>%
  select(country, year, hiv_prevalence) %>%
  pivot_wider(names_from = year, values_from = hiv_prevalence)
#
datatable(table1, rownames=F)
```

```


# Conclusions

`r if(current_continent == "Asia"){"This will only print if continent is Asia."}`

`r if(current_continent == "Americas"){"This will only print if continent is Americas."}`

`r if(current_continent == "Europe"){"This will only print if continent is Europe."}`

`r if(current_continent == "Africa"){"This will only print if continent is Africa."}`


